---
title: "Geospatial Data Wrangling with R"
editor: visual
---

## 1. Getting Started

The code chunk below will install and load tidyverse and sf packages.

```{r}
pacman::p_load(sf, tidyverse)
```

## 2. Data Acquisition

### 1.1 Importing polygon feature data in shapefile format

This code chunk will import ESRI shapefile into R.

```{r}
mpsz = st_read(dsn = "Hands-on_Ex1/data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### 1.2 Importing polyline feature data in shapefile format

```{r}
cyclingpath = st_read(dsn = "Hands-on_Ex1/data/geospatial", 
                         layer = "CyclingPathGazette")
```

### 1.3 Importing GIS data in kml format

```{r}
preschool = st_read("Hands-on_Ex1/data/geospatial/pre-schools-location-kml.kml")
```

## 2. Checking the Content of A Simple Feature Data Frame

### 2.1 Working with st_geometry()

```{r}
st_geometry(mpsz)
```

### 2.2 Working with glimpse()

```{r}
glimpse(mpsz)
```

### 2.3 Working with head()

```{r}
head(mpsz, n=5)
```

## 3. Plotting the Geospatial data

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

## 4. Working with Projection

### 4.1 Assigning EPSG code to a simple feature data frame

```{r}
st_crs(mpsz)
```

### 4.2 Assigning correct EPSG code to mpsz dataframe

```{r}
mpsz3414 <- st_transform(mpsz, 3414)
```

```{r}
st_crs(mpsz3414)
```

### 4.3 Transforming the projection of preschool from wgs84 to svy21

```{r}
st_geometry(preschool)
```

### 4.4 Performing project transformation using code change

```{r}
preschool3414 <- st_transform(preschool, 
                              crs = 3414)
```

```{r}
st_geometry(preschool3414)
```

## 5. Importing and converting An Aspatial Data

### 5.1 Importing the aspatial data

```{r}
listings <- read_csv("Hands-on_Ex1/data/aspatial/listings.csv")
```

```{r}
list(listings) 
```

### 5.2 Creating a simple feature dataframe from an aspatial dataframe

```{r}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

## 6. Geoprocessing with sf package

### 6.1 Buffering

```{r}
buffer_cycling <- st_buffer(cyclingpath, 
                               dist=5, nQuadSegs = 30)
```

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

```{r}
sum(buffer_cycling$AREA)
```

### 6.2 Point-in-polygon count

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))
```

```{r}
summary(mpsz3414$`PreSch Count`)
```

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
```

```{r}
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

## 7. Exploratory Data Analysis (EDA)

### 7.1 Histogram

```{r}
hist(mpsz3414$`PreSch Density`)
```

```{r}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

### 7.2 Scatterplot

```{r}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`), y= as.numeric(`PreSch Count`)))+
  geom_point() +
  labs(x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```
